---
title: "Urban Fires | Content Validation"
author: "Regina Bispo"
date: "2023-02-07"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3    
    number_sections: true
---

Run Raw2Final.Rmd before running this file.

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  tidy = TRUE,
	eval = TRUE,
	message = FALSE,
	warning = FALSE,
	comment = " ",
	include = TRUE
)

library(ggplot2)
library(dplyr)
library(maps)
library(dbscan)
library(factoextra)
library(RgoogleMaps)
library(maps) 
library(mapdata) 
library(rgdal)
library(viridis)
library(RColorBrewer)
library(tidygeocoder)
library(dplyr)
library(raster)
library(tmap)
library(rgeos)
library(sf)
library(spatstat)
library(geosphere)
library(readxl)
library(googleway)
library(deldir)
library(OpenStreetMap)
library(hms)
library(janitor)
library(stringr)
library(stringi)
library(here)
library(stargazer)
library(writexl)
library(formatR)
library(xlsx)
library(knitr)
library(lubridate)
library(validate)
library(gridExtra)


load('Raw2FinalEnv.RData')
```

# Data description

```{r}
summary(data)
```

## Variable length

+ Minutes:

```{r}
summary(data$length)
```

+ Days:

```{r}
summary(data$length/(24*60))
```

+ Years:

```{r}
summary(data$length/(24*60*365))
```

The next chunk provides the variable empirical quantiles:

```{r}
qtab("length")
```

The next chunk excludes cases with a fire length above 510 minutes (0.99 quantile):

```{r}
data = data %>%
  filter(data$length<=quantile(data$length,0.99))
```

Checking correction:

```{r}
summary(data$length)
```


```{r,include=FALSE}
hist(data$length,probability=T,main=" ",xlab="Length of urban fires (in minutes)")
```


```{r,include=FALSE}
n9=dim(data);n9
```

At this step, the dataset included `r n9[1]` objects and `r n9[2]` variables.


## Variable other

```{r}
summary(data$other)
```

The next chunk excludes cases with this variable above .

```{r}
data = data %>%
  filter(data$other<=quantile(data$other,0.99))
```

Checking correction:

```{r}
summary(data$other)
```


```{r,include=FALSE}
n10=dim(data);n10
n9-n10
```

At this step, the dataset included `r n10[1]` objects and `r n10[2]` variables.


## Variable otherv

```{r}
summary(data$otherv)
```

The next chunk excludes cases with a fire length above 0.99 quantile:

```{r}
data = data %>%
  filter(data$otherv<=3)
```

Cheching correction:

```{r}
summary(data$otherv)
```

```{r,include=FALSE}
n11=dim(data);n11
n10-n11
```

At this step, the dataset included `r n11[1]` objects and `r n11[2]` variables.


# Updating and checking rules

```{r}
new.rules <- validator(
  rule.id.1=is.numeric(data$id),
  rule.id.2=is_unique(data$id),
  rule.importance.1=is.factor(data$importance),
  rule.importance.2=data$importance %in% c("low","median","high","unknown"),
  rule.importance.cv.1= with(data,sum((deaths==0 & (major+minor)<5 & length<30)))==sum(data$importance=="low") ,
  rule.importance.cv.2= with(data,sum((deaths>=1 | (major+minor)>=10 | length>=60)))==sum(data$importance=="high"),
  rule.importance.cv.3= with(data,sum(deaths==0 & ((major+minor)>=5 & (major+minor)<10) | (length>=30 & length<60)&length<60))==sum(data$importance=="medium"),
  rule.open.1=is.POSIXct(data$open),
  rule.open.2=in_range(data$open,min="2013-01-01 00:00:00",max="2022-12-31 23:59:00"),
  rule.close.1=is.POSIXct(data$close),
  rule.close.2=in_range(data$close,min="2013-01-01 00:00:00",max="2023-01-01 23:59:00"),
  rule.close.3=data$close > data$open,
  rule.length.1=is.numeric(data$length),
  rule.length.2=data$length == difftime(data$close,data$open,units = "mins"),
  rule.length.cv=data$length<=501,
  rule.type.1=is.factor(data$type),
  rule.type.2=length(levels(data$type))==length(structure.list),
  rule.type.3=data$type %in% structure.list,
  rule.region.1=is.factor(data$region),
  rule.region.2=length(levels(data$region))==length(region.list),
  rule.region.3=data$region %in% region.list,
  rule.subregion.1=is.factor(data$subregion),
  rule.subregion.2=length(levels(data$subregion))==length(subregion.list),
  rule.subregion.3=data$subregion %in% subregion.list,
  rule.district.1=is.factor(data$district),
  rule.district.2=length(levels(data$district)) <= length(levels(as.factor(dist$NAME_1))),
  rule.district.3=data$district %in% dist$NAME_1,
  rule.municipality.1=is.factor(data$municipality),
  rule.municipality.2=length(levels(data$municipality)) <= length(levels(as.factor(conc$NAME_2))),  
  rule.municipality.3=data$municipality %in% levels(as.factor(conc$NAME_2)),  
  rule.parish.1=is.factor(data$parish),
  rule.parish.2=length(levels(data$parish)) <= length(levels(as.factor(freg$NAME_3))),
  rule.lat.1=is.numeric(data$lat),
  rule.lat.2=in_range(data$lat, min=dist@bbox[2,1], max=dist@bbox[2,2]),
  rule.lon.1=is.numeric(data$lon),
  rule.lon.2=in_range(data$lon, min=dist@bbox[1,1], max=dist@bbox[1,2]),
  rule.pfd.1=is.factor(data$pfd),
  rule.groundhr.1=is.numeric(data$groundhr),
  rule.groundhr.2=data$groundhr >= 0,
  rule.groundtr.1=is.numeric(data$groundtr),
  rule.groundtr.2=data$groundtr >= 0,
  rule.airhr.1=is.numeric(data$airhr),
  rule.airhr.2=data$airhr >= 0,
  rule.airtr.1=is.numeric(data$airtr),
  rule.airtr.2=airtr >= 0,
  rule.deaths.1=is.numeric(data$deaths),
  rule.deaths.2=data$deaths >= 0,
  rule.major.1=is.numeric(data$major),
  rule.major.2=data$major >= 0,
  rule.minor.1=is.numeric(data$minor),
  rule.minor.2=data$minor >= 0,
  rule.assist.1=is.numeric(data$assist),
  rule.assist.2=data$assist >= 0,
  rule.other.1=is.numeric(data$other),
  rule.other.2=data$other >= 0,
  rule.other.cv=data$other <=3,
  rule.anepcv.1=is.numeric(data$anepcv),
  rule.anepcv.2=data$anepcv >= 0,
  rule.otherv.1=is.numeric(data$otherv),
  rule.otherv.2=data$otherv >= 0,
  rule.otherv.cv=data$otherv <=3,
  rule.reignit.1=is.factor(data$reignit),
  rule.reignit.2=length(levels(data$reignit))==2,
  rule.reignit.3=data$reignit %in% c("no","yes")
)


out=confront(data, new.rules)
summary(out)[-c(6:8)]


```

## Exploratory data analysis

## Fire importance

The next chunk gives the distribution of the fire importance (table and plot):

```{r}
ctab(var = importance)
```

## Ignition date/time

The next chunks gives the distribution of the events as a function of:


+ Year

```{r}
ctab(var = format(open,format="%Y"))[[1]]
```


+ Month of occurence

```{r}
ctab(var = lubridate::month(data$open,label = TRUE))
```

+ Weekday

```{r}
tab=data %>% 
  count(wd=factor(weekdays(data$open),
                  levels = unique(weekdays(data$open)),
                  labels = c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday")), sort = TRUE) %>% 
  arrange(desc(n)) %>%
  mutate(prop=n/nrow(data)) %>% arrange(-prop)

ggplot(tab, aes(y = reorder(wd, prop),x=prop)) +
 geom_bar(stat="identity", position = position_stack(reverse = TRUE),fill="black",alpha=0.7) + ylab(" ") + xlab("Proportion") +
 theme(legend.position = "top")+theme_bw()
```


+ Hour of the day

```{r}
ctab(var = as.POSIXlt(data$open)$hour)
```


```{r,fig.width=6,fig.height=5,include=FALSE}
v1=format(data$open,format="%Y")
v2=factor(lubridate::month(data$open,label = TRUE),levels = unique(month(data$open,label = TRUE)), labels = c("jan","feb","mar","apr","may","jun","jul","aug","sep","oct","nov","dec"))
v3=factor(weekdays(data$open), levels = unique(weekdays(data$open)), labels = c("Sunday","Monday","Tuesday","Wednesday","Thursday","Friday","Saturday")) 
v4=as.POSIXlt(data$open)$hour
  
tab1=data %>%
  mutate(x=v1) %>%
  count(x, sort = TRUE) %>% 
  mutate(Proportion=round(n/nrow(data),4)) %>% arrange(-Proportion) 

p1=ggplot(tab1, aes(x,Proportion)) + 
  geom_bar(stat = "identity")+
  labs(x=" ",y="Frequency")+theme_bw()+
  geom_bar(stat="identity",color="black",fill="white")+  
  geom_text(aes(label = paste0(sprintf("%4.1f", Proportion * 100), "%")),cex=3,vjust=2)+ ggtitle("(A)")

tab2=data %>%
  mutate(x=v2) %>%
  count(x, sort = TRUE) %>% 
  mutate(Proportion=round(n/nrow(data),4)) %>% arrange(-Proportion) 

p2=ggplot(tab2, aes(x,Proportion)) + 
  geom_bar(stat = "identity")+
  labs(x=" ",y="Frequency")+theme_bw()+
  geom_bar(stat="identity",color="black",fill="white")+  
  geom_text(aes(label = paste0(sprintf("%4.1f", Proportion * 100), "%")),cex=3,vjust=2)+ ggtitle("(B)")

tab3=data %>%
  mutate(x=v3) %>%
  count(x, sort = TRUE) %>% 
  mutate(Proportion=round(n/nrow(data),4)) %>% arrange(-Proportion) 

p3=ggplot(tab3, aes(x,Proportion)) + 
  geom_bar(stat = "identity")+
  labs(x=" ",y="Frequency")+theme_bw()+
  geom_bar(stat="identity",color="black",fill="white")+  
  geom_text(aes(label = paste0(sprintf("%4.1f", Proportion * 100), "%")),cex=3,vjust=2)+ ggtitle("(C)")

tab4=data %>%
  mutate(x=v4) %>%
  count(x, sort = TRUE) %>% 
  mutate(Proportion=round(n/nrow(data),4)) %>% arrange(-Proportion) 

p4=ggplot(tab4, aes(x,Proportion)) + 
  geom_bar(stat = "identity")+
  labs(x=" ",y="Frequency")+
  geom_bar(stat="identity",color="black",fill="white")+  
  geom_text(aes(label = paste0(sprintf("%4.1f", Proportion * 100), "%"),angle=90),cex=3,hjust=1.2) +
  theme_bw()+ ggtitle("(D)")

grid.arrange(p1,p2,p3,p4, nrow = 2)
```

## Total counts per district

```{r maps,fig.width=10,fig.height=5,warning=FALSE}
#Fire data by district
map=dist[dist$NAME_0=="Portugal",]
map$NAME_1=factor(map$NAME_1)
PTdist=data %>% count(district, sort = TRUE)
colnames(PTdist)=c("NAME_1","n")
levels(PTdist$NAME_1)=levels(map$NAME_1)
mapadist=merge(map,PTdist,by="NAME_1", all.x=T)

#Fire departments by district
FSdist=data %>%
  group_by(district) %>%
  summarize(count_distinct = n_distinct(pfd))
colnames(FSdist)=c("NAME_1","n")
levels(FSdist$NAME_1)=levels(map$NAME_1)
mapadistfs=merge(map,FSdist,by="NAME_1", all.x=T)
```

Mapping:

```{r}
p1=tm_shape(mapadist)+
  tm_polygons("n", style="kmeans",palette="Reds",breaks=quantile(mapadist$n),title="N. of events")+
  tm_text("NAME_1",size = 0.5, fontfamily="sans", ymod = 0.4)+
  tm_graticules()+
  tm_layout(legend.outside = TRUE, legend.outside.position = "right")+
tm_shape(mapadist)+
  tm_polygons("n", style="kmeans",palette="Reds",breaks=quantile(mapadist$n),legend.show=FALSE,alpha = 0.1)+
  tm_text("n", size = 0.5, fontfamily="sans", ymod = -.1)

bks=c(11,21,26,31,41,56)
p2=tm_shape(mapadistfs)+
  tm_polygons("n",palette="Greens",breaks=bks,title="N. of fire departments")+
  tm_text("NAME_1",size = 0.5, fontfamily="sans", ymod = 0.4)+
  tm_graticules()+
  tm_layout(legend.outside = TRUE, legend.outside.position = "right")+
tm_shape(mapadistfs)+
  tm_polygons("n", palette="Greens",breaks=bks,legend.show=FALSE,alpha = 0.1)+
  tm_text("n", size = 0.5, fontfamily="sans", ymod = -.1)


tmap_arrange(p1,p2)
```

## Variable type


```{r}
tab=data %>% 
  count(type, sort = TRUE) %>% 
  mutate(prop=n/nrow(data)) %>% arrange(-prop) 

ggplot(tab, aes(y = reorder(type, prop),x=prop)) +
 geom_bar(stat="identity", position = position_stack(reverse = TRUE),color="black",fill="white") + ylab(" ") + xlab("Proportion") +
 theme(legend.position = "top")+theme_bw()+ 
  geom_text(aes(label = paste0(sprintf("%4.2f", prop * 100), "%")),cex=3,hjust=0.01,nudge_x = 0.01) +
  theme(plot.margin = unit(c(2,3,2,2), "cm"))+xlim(c(0,1))+ ggtitle("(A)")
```

### Total counts per type (residential vs. not)

```{r maps1,fig.width=10,fig.height=5,warning=FALSE}
#Residential
TypeDist=data[data$type=="Residential",] %>%
  count(district, sort = TRUE)
colnames(TypeDist)=c("NAME_1","n")
levels(TypeDist$NAME_1)=levels(map$NAME_1)
mapadisttype=merge(map,TypeDist,by="NAME_1", all.x=T)

# Other
TypeDisto=data[data$type!="Residential",] %>%
  count(district, sort = TRUE)
colnames(TypeDisto)=c("NAME_1","n")
levels(TypeDisto$NAME_1)=levels(map$NAME_1)
mapadisttypeo=merge(map,TypeDisto,by="NAME_1", all.x=T)

tmap_options(check.and.fix = TRUE)

p3=tm_shape(mapadisttype)+
  tm_polygons("n",palette="Blues",breaks=quantile(mapadisttype$n),title="N. of events")+
  tm_text("NAME_1",size = 0.5, fontfamily="sans", ymod = 0.4)+
  tm_graticules()+
  tm_layout(legend.outside = TRUE, legend.outside.position = "right")+#,title = "(B)"
tm_shape(mapadisttype)+
  tm_polygons("n",palette="Blues",breaks=quantile(mapadisttype$n),legend.show=FALSE,alpha = 0.1)+
  tm_text("n", size = 0.5, fontfamily="sans", ymod = -.1)


p4=tm_shape(mapadisttypeo)+
  tm_polygons("n", palette="Blues",breaks=quantile(mapadisttype$n),title="N. of events")+
  tm_text("NAME_1",size = 0.5, fontfamily="sans", ymod = 0.4)+
  tm_graticules()+
  tm_layout(legend.outside = TRUE, legend.outside.position = "right")+#,title = "(C)"
 tm_shape(mapadisttypeo)+
   tm_polygons("n", palette="Blues",breaks=quantile(mapadisttype$n),legend.show=FALSE,alpha = 0.1)+
  tm_text("n", size = 0.5, fontfamily="sans", ymod = -.1)

tmap_arrange(p3,p4)
```

## Frequency of fires per location

The next chunk builds a dataframe quantifying the number of objects with exactly the same coordinates. 

```{r}
namesd=sort(unique(data$district))
years=as.numeric(unique(format(data$open,format="%Y")))

uflocal<-list()

for(i in namesd){
  uflocalyear<-data.frame()
  for (j in 1:length(years)) {
  temp<-nburn(i,years[j])
  uflocalyear<-rbind(uflocalyear,temp)
  }
  uflocal[[i]]<-uflocalyear
  }

ufreps<-bind_rows(Filter(is.data.frame, uflocal))
rownames(ufreps)<-NULL
```

This chunk presents an example:

```{r}
ufreps[ufreps$District=="Coimbra"&ufreps$Year==2022,]

```


In this example there were `r ufreps[ufreps$District=="Coimbra",][3,3]` locations for which only `r ufreps[ufreps$District=="Coimbra",][3,2]` event was registered.


The next chunk adds a new variable (cty) to the dataset to distinguish this type of coordinates

```{r}
#spatial coordinates
crep=data.frame(data[,c("district","open","lon","lat")])
crep2sp<-data.frame(data[,c("district","open","lon","lat")])
crep2sp$year = as.numeric(format(data$open,format="%Y"))
coordinates(crep2sp)=~lon+lat
#builds a list containing points with zero distance by year
list.d0=sapply(split(crep2sp, as.numeric(format(data$open,format="%Y"))), zerodist)
#transforms to data.frame
df.d0=as.data.frame(do.call(rbind, unname(Map(cbind, id = names(list.d0), list.d0))))
colnames(df.d0)=c("year","v1","v2")

v1<-vector()
#new variable by year --- cty --- Coordinates TYpe
for (i in 1:length(years)) {
v0=ifelse(data$lon[format(data$open,format="%Y")==years[i]] %in% data$lon[as.numeric(df.d0[df.d0$year==years[i],2])],"not exact","exact")
v1=c(v1,v0)
}
data$cty<-factor(v1)
```

The next chunk produces a bubble plot for each one of the districts, in which the size of the bubble is proportion to the num of fires within the same location.

```{r,fig.width=18,fig.height=8}
# Bubble plot
crep$year<-format(crep$open,format="%Y")
bubble=crep[crep$lon %in% crep[df.d0[,2],"lon"],] %>%
    arrange(district,year,lon) %>%
  count(district,year,lon,lat)
coordinates(bubble)=~lon+lat
bks=c(2,4,6,Inf)
tmap_options(check.and.fix = TRUE)
for (i in 1:length(unique(bubble$district))) {
  name=paste(unique(bubble$district)[i])
 tm_shape(freg[freg$NAME_1==name,]) + 
          tm_borders() + 
          tm_dots(size=0.1,col="red") + 
        tm_shape(bubble[bubble$district==name,])+ 
          tm_bubbles(size = "n", col="n",alpha=0.4, breaks=bks, legend.size.show = FALSE, title.col = "Frequency") +  tm_facets(by = "year", ncol = 6) + 
          tm_layout(main.title= paste(name), main.title.position = "center",main.title.size = 3,legend.text.size = 2, legend.title.size = 3,frame=FALSE,legend.outside.size = 0.3,panel.label.size = 3)
  }

```

Example:

```{r}
i=3
name=paste(unique(bubble$district)[i])
  print(tm_shape(freg[freg$NAME_1==name,]) + 
          tm_borders() + 
          tm_dots(size=0.03,col="red") + 
        tm_shape(bubble[bubble$district==name&bubble$year=="2013",])+ 
          tm_bubbles(size = "n", col="n",alpha=0.4, breaks=bks, legend.size.show = FALSE, title.col = "Frequency") + 
          tm_layout(main.title= " ", main.title.position = "center",main.title.size = 2,legend.text.size = 1, legend.title.size = 1.5,frame=FALSE))
```

Variable distribution:
```{r}
ctab(var = cty)
```

Description by distric:
```{r}
data %>%
  group_by(district) %>%
  count(cty) %>%
  group_by(district) %>%
  mutate(perc=round(n/sum(n)*100,1))
```



## Administrative units

#### District

```{r}
tab=data %>% 
  count(district, sort = TRUE) %>% 
  arrange(desc(n)) %>%
  mutate(prop=n/nrow(data)) %>% arrange(-prop)

tab

ggplot(tab, aes(y = reorder(district, prop),x=prop)) +
 geom_bar(stat="identity", position = position_stack(reverse = TRUE),fill="black",alpha=0.7) + ylab(" ") + xlab("Proportion") +
 theme(legend.position = "top")+theme_bw()
```

#### Municipality

```{r}
tab=data %>% 
  group_by(district) %>%
  count(municipality, sort = TRUE) %>% 
  arrange(desc(n)) %>%
  mutate(prop=round(n/nrow(data),4)) %>% arrange(-prop)
tab[tab$prop>0.01,]
```


## Resources

### Ground human resources

```{r}
qtab(var = "groundhr")

table(cut(data$groundhr,c(0,5,10,15,20,25,30,141)))
prop.table(table(cut(data$groundhr,c(0,5,10,15,20,25,30,141))))*100

```


### Ground tecnhical resources

```{r}
qtab(var = "groundtr")

plot(table(data$groundtr))
table(cut(data$groundtr,c(0,1,3,5,7,9,11,49)))
prop.table(table(cut(data$groundtr,c(0,1,3,5,7,9,11,49))))*100

```

### Air resources

+ human

```{r}
table(data$airhr)
round(prop.table(table(data$airhr)),3)
```

+ technical

```{r}
table(data$airtr)
round(prop.table(table(data$airtr)),3)
```

## Victims

+ major injuries

```{r}
table(data$major)
round(prop.table(table(data$major)),3)

```

+ minor injuries

```{r}
table(data$minor)
round(prop.table(table(data$minor)),3)

```

+ deaths

```{r}
table(data$deaths)
round(prop.table(table(data$deaths)),3)

```

## Reignition

```{r}
ctab(var = reignit)
```
